**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-09 Manuscripts/Manuscript_AM_v1.Rmd
Updated
Updated background, references

A longitudinal study is defined as one where a variable of interest is measured repeatedly in a group (or groups) of subjects. In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment across time, rather than analyzing the information at a single timepoint (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening measured was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data, a statistical view that derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008]. when applied to longitudinal, a frequentist analysis is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: a constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004]. The first condition is frequently unjustified  as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016]. The reasons behind the second assumption being not met in biomedical research are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a certain timepoint. When these issues arise, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can lead to increased costs for the study if the desired statistical power is not met with the remaining subjects, and can lead to the exclusion of valuable information.Additionaly, rm-ANOVA uses a _post hoc_ analysis to assess significance in the differences between groups. Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].

Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010]. Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006]. These models are more flexible than rm-ANOVA as they can accommodate missing observations in the data, and  allow different modeling strategies for the covariance[@west2014]. On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].

The assumption of linearity is a key component of LMEMs and rm-ANOVA, and therefore both models restrict the inferences they can extract from longitudinal data when it does not follow a linear trend. In biomedical research, such behavior in longitudinal data does arise and has been exemplified in different studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. In such situations, an estimated model via rm-ANOVA or LMEMs may not provide a consistent fit with the observed variations in the variables of interest. Moreover, although it is possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived from it.

Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to use statistical tools that allow the data to determine the fit of the model while enabling  inferences that are  both adequate and consistent from a statistical standpoint.

@@

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017]. Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models for longitudinal data.

Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective. First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis. Secondly, it uses simulation to generate non-linear longitudinal data and presents the implementation of GAMs as a statistical tool that enables the analysis of data that shows this behavior. And finally, it introduces Bayesian statistics and presents their implementation with GAMs over simulated data to demonstrate the differences and benefits of this approach . With an emphasis on reproducibility by providing the code and dataset used, this study provides biomedical researchers with a clear view of the advantages of different approaches in the statistical analysis of longitudinal data.

@@

### 1 The _frequentist_ case for longitudinal data

While the _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, there are key assumptions that are made in order to make the model valid. From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects. Each one of these assumptions is discussed below.

@@

In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects. The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].

>>>>

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from the results of a longitudinal study. Such statistical view derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: a constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study  [@schober2018;@gueorguieva2004].

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases[@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations. In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.

When the aforementioned issues arise, rm-ANOVA requires to exclude all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information. This is exemplified by the pioneering use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].

Additionally, another assumption is made in both rm-ANOVA and LMEMs models, where a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. This studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.

@@

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].

Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of GAMs and Bayesian models for longitudinal data.In particular, $\textsf{R}$ has an extensive collection of documentation, functions and libraries that speed up the initial stages of the analsys and that enable the use of complex statistical methods without requiring a specialized set of programming skills from the user.

Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective. First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis. Secondly, it uses $\textsf{R}$ to simulate non-linear longitudinal data following previously reported values in the literature, and presents the implementation of GAMs as a statistical tool for longitudinal data. And finally, it introduces Bayesian statistics and presents their implementation with GAMs to demonstrate the differences and benefits of this approach. with an emphasis in reproducibility  by providing the code, simulated dataset and a step-by-step guide of the computational implementation of different models, this study aims to encourage the exploration of modern statistical methods for biomedical longitudinal data, and to improve the  statistical standards in biomedical research.

@@

### 1 The "frequentist" case for longitudinal data

The _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, and there are key assumptions that are made in order to make the model valid. From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects. Each one of these assumptions is discussed below.

@@

In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects. The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002]. The model then, is a linear combination of terms, and if plotted, it would a straight line.

#### 1.2 Constant correlation
